---
title: "Eight hard problems in criminology, with proofs behind the functions"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Eight hard problems in criminology, with proofs behind the functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rmorie)
```

Every function on this page rests on a theorem checked in Lean 4 with
Mathlib (the sources ship under `research/lean/`; `Audit.lean` prints the
axioms, which are the three standard ones, and no file contains `sorry`).
The rule throughout: Lean certifies the implication, never the
antecedent. Each section says where the empirical assumption enters.

# P4. Predictive-policing feedback loops run away, and slowly

Two regions, true rates 0.3 and 0.2, patrol allocated in proportion to
discovered crime. The mean-field recursion converges to sending the
whole patrol to the higher-rate region, whatever the ratio of the rates
(`Research.P4.naiveShare_tendsto_one`):

```{r p4}
mf <- morie_feedback_loop_meanfield(0.3, 0.2, 10, 10, n_steps = 20000)
tail(mf$share_a, 1)
attr(mf, "limit")$theorem
```

The corrected update of Ensign et al. (discount each discovery by the
presence that produced it) converges to the true-rate proportion instead
(`Research.P4.corrected_share_tendsto`):

```{r p4c}
cf <- morie_feedback_loop_meanfield(0.3, 0.2, 10, 10, n_steps = 20000, update = "corrected")
tail(cf$share_a, 1)
```

The limit is not the whole story. The proved rate bound
(`Research.P4.naiveShare_rate_bound`) shows the share can move at most
like a harmonic sum: a 5 percent rate gap from a 1/99 start cannot move
the allocation by more than about a tenth in two million steps.

```{r p4b}
morie_feedback_loop_bound(0.21, 0.20, 1, 99, n_steps = 2e6)$bound
```

Assumption that enters: that discovery is proportional to presence
times the true rate.

# P5. What a risk score's fairness can and cannot be

Chouldechova's identity ties the false positive rate to base rate, PPV
and FNR (`Research.P5.Table.chouldechova`); two groups with equal PPV and
FNR but different base rates cannot share an FPR
(`Research.P5.impossibility`):

```{r p5}
morie_fairness_implied_fpr(p = 0.30, ppv = 0.6, fnr = 0.25)
morie_fairness_implied_fpr(p = 0.50, ppv = 0.6, fnr = 0.25)
```

With rearrest as a noisy proxy for reoffending, the true base rate is
identified only up to an interval, and both ends are attainable
(`Research.P5.true_base_rate_bounds`); whether two groups can even be
ordered depends on whether the intervals overlap
(`Research.P5.compare_decided`, `compare_undecided`):

```{r p5b}
morie_fairness_base_rate_bounds(c(0.35, 0.55), alpha_max = 0.05, beta_max = 0.20)
morie_fairness_compare_groups(0.35, 0.55, alpha_max = 0.05, beta_max = 0.40)[c("order", "breakdown_beta_max")]
```

Assumption that enters: the noise boxes.

# P1. The dark figure as an interval, not a point

Two lists of the same events identify the true count only under
independence; with a dependence factor in a box the count is an
interval with attainable ends (`Research.P1.TwoSource.petersen_bounds`):

```{r p1}
morie_dark_figure_two_source(400, 250, 80)
morie_dark_figure_two_source(400, 250, 80, kappa = 2)[c("lower", "upper")]
```

A recorded rate and a survey rate together bound the true rate and the
dark figure (`Research.P1.dark_figure_bounds`), and the breakdown value
says how much under-reporting would overturn a conclusion
(`Research.P1.conclusion_holds_below_breakdown`):

```{r p1b}
morie_dark_figure_bounds(v_obs = 0.06, r = 0.02, alpha_max = 0.01, beta_max = 0.30)
morie_dark_figure_breakdown(0.06, threshold = 0.10)$breakdown_beta_max
```

# P3. Hot-spot effects when places affect their neighbours

An exposure mapping on a place network (treated / untreated with a
treated neighbour / isolated), the stratified exposure adjustment
(`Research.P3.Model.exposure_adjustment`), the direct/spillover split, and
the bias of pooling exposures 0 and 1 as "control":

```{r p3}
set.seed(1)
n <- 400
stratum <- rep(c("core", "edge"), each = n / 2)
exposure <- sample(0:2, n, replace = TRUE)
y <- 5 - 1.0 * (exposure == 2) - 0.4 * (exposure == 1) + (stratum == "core")
morie_spillover_effects(y, exposure, stratum)[c("spillover", "direct", "total")]
```

Under a randomised deployment the Horvitz–Thompson estimator needs no
ignorability at all (`Research.P3.Design.ht_unbiased`):

```{r p3ht}
edges <- cbind(1:7, 2:8)
pr <- morie_spillover_exposure_probs(n = 8, edges, n_treated = 3)
trt <- c(2, 5, 7)
e <- morie_spillover_exposure(seq_len(8) %in% trt, edges)$exposure
morie_spillover_ht(y = 10 + seq_len(8) - 3 * (e == 2) - (e == 1), e, pr)$total_effect
```

# P7. How much of the "law of crime concentration" is crime-free places?

The Gini of all places splits exactly into the zero share and the Gini
among places with any crime (`Research.P7.gini_zero_decomposition`), and
a uniform Poisson process produces a zero share of exactly `exp(-mu)`
(`Research.P7.poisson_zero_prob`):

```{r p7}
set.seed(1)
x <- rpois(2000, 0.4)
unlist(morie_concentration_decompose(x)[c("zero_share", "null_zero_share", "gini_all", "gini_positive")])
```

# P2, P6, P8 in one breath

Disparity with exposure known only within a factor (`Research.P2.disparity_bounds`),
the aggregate age–crime curve of a mixture and its identical one-type
twin (`Research.P6.aggregate_not_identifying`), and which deterrence
partial effects a design can identify (`Research.P8.constant_dimension_not_identified`):

```{r p268}
morie_disparity_exposure_bounds(y = c(A = 300, B = 100), m = c(A = 1000, B = 1000), gamma = 1.5, reference = "B")[, c("group", "ratio_lower", "ratio_upper", "direction_identified")]
ages <- 10:60
early <- dgamma(ages - 9, shape = 3, rate = 0.5) * 40
late <- dgamma(ages - 9, shape = 8, rate = 0.35) * 40
two <- morie_age_crime_aggregate(c(0.5, 0.5), cbind(early, late), ages)
head(attr(two, "equivalent_single_type"), 3)
morie_deterrence_design_check(p = c(0.1, 0.3, 0.5), s = c(2, 2, 2), c = c(30, 30, 30))$identified
```

# References

Ensign D, Friedler SA, Neville S, Scheidegger C, Venkatasubramanian S (2018).
Runaway feedback loops in predictive policing. PMLR 81, 160–171.

Chouldechova A (2017). Fair prediction with disparate impact. Big Data 5(2), 153–163.

Bird SM, King R (2018). Multiple systems estimation to inform public policy.
Annual Review of Statistics and Its Application 5, 95–118.

Weisburd D (2015). The law of crime concentration and the criminology of place.
Criminology 53(2), 133–157.

Nagin DS (2013). Deterrence in the twenty-first century. Crime and Justice 42, 199–263.
