---
title: "Designing and measuring for a small subpopulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing and measuring for a small subpopulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What this vignette is for

Research on a subpopulation that is a small share of the general frame runs
into three problems at once, and they compound. The group is small, so a
general sample yields few of its members. It is often reached through
clustered or otherwise non-random designs, so the effective sample is smaller
than the nominal one. And the frame may identify the group imperfectly, so
even the members who are sampled may not be counted as such.

The functions here address all three. They are written as general survey
methodology; the running application is estimation for Indigenous populations
in Canada, where all three conditions hold together.

## 1. How large must the sample be?

The naive calculation is the textbook one for a proportion:

```{r}
library(rmorie)
morie_sample_size_proportion(p = 0.5, moe = 0.05)
```

which returns about 385. That number is almost never the answer for
subpopulation work, because it is the number of *subgroup members* needed,
not the number of interviews. If the group is 5% of the frame and members are
selected at the same rate as everyone else, the general sample has to be
roughly twenty times larger:

```{r}
morie_sample_size_domain(p = 0.5, moe = 0.05, domain_prevalence = 0.05)
```

The `n_overall` element is near 7,700. This is the single most common reason
that general-population surveys cannot report on small subpopulations: they
were never sized to.

Three further factors move the requirement, and all of them move it upward.

```{r}
morie_sample_size_domain(
  p = 0.5, moe = 0.05,
  domain_prevalence = 0.05,
  deff = 1.8,            # clustering by community and unequal weights
  response_rate = 0.55,  # realistic fieldwork
  coverage = 0.82        # the frame misses some members
)
```

The design effect multiplies the requirement, the response rate divides it,
and the coverage rate divides it again. `deff` can be built from its parts:
`morie_deff_cluster()` for the clustering component and
`morie_design_effect()` for the weight component. They multiply.

```{r}
morie_deff_cluster(m = 25, icc = 0.02)   # 25 interviews per community
morie_neff_cluster(n = 500, m = 25, icc = 0.02)
```

## 2. Reaching the group: oversampling and screening

Two designs raise subgroup yield. The first is disproportionate selection:
sample the group at a higher rate than the rest of the frame.

```{r}
morie_oversample_factor(domain_prevalence = 0.05, target_share = 0.20)
```

`factor` is how much higher the selection probability must be. Note
`deff_weights`: oversampling is not free. It creates unequal weights, which
inflate the design effect, which raises the sample size requirement. Plan the
two together rather than in sequence.

The second design is screening, used when the frame does not identify the
group at all.

```{r}
morie_screen_design(
  prevalence = 0.05, target_n = 400,
  cost_screen = 5, cost_interview = 120,
  screen_response = 0.7, interview_response = 0.85
)
```

`screening_share_of_cost` is the quantity to look at when deciding whether
screening is viable: for a rare group most of the budget is spent finding
people rather than interviewing them.

When strata differ in cost, as remote communities usually do, allocation
should account for it:

```{r}
morie_alloc_optimal(
  N_h = c(8000, 1500, 500), S_h = c(1, 1.4, 2),
  n = 900, cost_h = c(1, 3, 9)
)
```

With equal costs this reduces to Neyman allocation, also available directly
as `Neyman()`.

## 3. Making the sample agree with known totals

Where population totals are published for the group but their joint
distribution is not, raking adjusts the weights to match each margin in turn:

```{r}
df <- data.frame(
  region = c("north", "north", "south", "south"),
  group  = c("a", "b", "a", "b")
)
morie_rake(df, list(region = c(north = 60, south = 40),
                    group  = c(a = 70, b = 30)))
```

Always check `converged`. Raking can fail to converge when margins are
inconsistent or a cell is empty, and a non-converged weight vector should not
be used.

## 4. When the identifier is wrong

If group membership is recorded by an administrative identifier rather than
by self-identification, the identifier has a sensitivity and a specificity,
and neither is 1. Sensitivity below 1 means true members are recorded as
non-members. Every rate computed on the recorded counts is then biased, and
usually biased downward, because the group's cases are undercounted while the
denominator is not.

```{r}
morie_misclass_correct(p_obs = 0.04, sensitivity = 0.75,
                       specificity = 0.999, n = 10000)
```

The correction is the Rogan-Gladen estimator. Applied to counts, it reports
how many members the records miss:

```{r}
morie_misclass_count(observed_count = 400, n = 10000,
                     sensitivity = 0.75, specificity = 0.999)
```

Two cautions. The correction needs external estimates of sensitivity and
specificity, normally from a linkage or validation study; it cannot be
obtained from the misclassified data alone. And the interval widens sharply
as the classifier degrades, which is itself the finding worth reporting when
no validation study exists.

## 5. Measurement: does the instrument mean the same thing in both groups?

Comparing scale scores across groups presupposes that the scale measures the
same construct in each. That is an empirical claim, and it is testable.

Plan the test first. Power for differential item functioning is governed by
the *smaller* group, so a small focal group caps detection regardless of the
total sample:

```{r}
morie_dif_sample_size(p_reference = 0.6, odds_ratio = 1.5)              # equal groups
morie_dif_sample_size(p_reference = 0.6, odds_ratio = 1.5, ratio = 4)   # focal is a fifth
```

An instrument reported as invariant on an underpowered comparison has not
been shown to be invariant; it has been shown to be untested. Report the
detectable effect size alongside the DIF result.

Fit the DIF models with `Difmh()` (Mantel-Haenszel) or `Difsib()` (SIBTEST),
then convert the odds ratios to the scale used for operational review:

```{r}
morie_dif_delta_mh(c(1.0, 1.35, 1.9, 0.5))
```

Magnitude A is negligible, B moderate, C large; `favours` names the group the
item is easier for. For scale-level invariance, compare the nested sequence:

```{r}
fits <- data.frame(
  model = c("configural", "metric", "scalar"),
  chisq = c(120.3, 128.9, 162.4),
  df    = c(48, 54, 60),
  cfi   = c(0.981, 0.979, 0.964),
  rmsea = c(0.041, 0.040, 0.052)
)
morie_invariance_compare(fits)
```

Metric invariance holds here and scalar invariance does not, which means
factor loadings are comparable but intercepts are not: group *means* on this
scale are not comparable, though associations may be. Reporting a mean
difference in that situation describes the instrument, not the groups.

Finally, report precision by group as well. A scale can be reliable in one
group and not in another, and pooling conceals it:

```{r}
morie_irt_theta_se(information = c(4, 9, 16))
morie_irt_marginal_reliability(se = c(0.3, 0.35, 0.4, 0.5))
```

## 6. Governance is part of the method

For research with Indigenous communities in Canada, the design questions
above sit inside a governance framework that is not optional and is not
supplied by any software package. First Nations data governance is
articulated through the OCAP principles -- ownership, control, access and
possession -- which are a registered trademark of the First Nations
Information Governance Centre and are set out by that body; Inuit and Metis
governance frameworks are distinct and are articulated by their own
representative organisations. A distinctions-based approach means First
Nations, Inuit and Metis data are not aggregated into a single category
without a reason that the relevant nations accept.

Practical consequences for the calculations above: who holds the data and
who decides what is published are determined before the sample is drawn, not
after the analysis; the community, not only the researcher, has a stake in
whether a subgroup estimate is precise enough to publish; and a
non-converged raking or an undercounted identifier is a reporting obligation,
not a technical footnote.

None of this is enforced by the functions in this package. It is stated here
because a design that is statistically sound and governance-blind is not
usable for this work.

## Summary

| Question | Function |
|---|---|
| How many subgroup members do I need? | `morie_sample_size_proportion()` |
| How large must the whole sample be? | `morie_sample_size_domain()` |
| What does clustering cost me? | `morie_deff_cluster()`, `morie_neff_cluster()` |
| How hard do I oversample, and at what price? | `morie_oversample_factor()` |
| Is screening affordable? | `morie_screen_design()` |
| How do I allocate across costly strata? | `morie_alloc_optimal()`, `Neyman()` |
| How do I match published totals? | `morie_rake()`, `Postrt()` |
| The identifier is imperfect | `morie_misclass_correct()`, `morie_misclass_count()` |
| Can I even detect DIF? | `morie_dif_sample_size()` |
| How large is the DIF? | `Difmh()`, `morie_dif_delta_mh()` |
| Are scores comparable across groups? | `morie_invariance_compare()` |
| How precise are the scores, by group? | `morie_irt_theta_se()`, `morie_irt_marginal_reliability()` |
